ODGI: understanding pangenome graphs

### 2021-10-30 manuscript/main.tex
italianization: implementation; explicit who is depth, use model only for the pangenomes, typos, little changes

Pangenome graphs can be constructed by multiple sequence alignment~\citep{Lee_2002,Grasso_2004} or by transitively reducing an alignment between sequences to an equivalent, labeled sequence graph~\citep{Kehr_2014,Garrison_2019_thesis}.
Current methods to build these graphs are still under active development~\citep{Li:2020,Armstrong:2020,pggb}, but they have largely settled on a common data model, represented in the Graphical Fragment Assembly (GFA) format \citep{GFA}).
%\FIXME{may want to move GFA standard+VG to intro section 1}. AG: it is borderline, but it seems fine to have it here too.
This standardization supports the development of a reference set of tools that operate on the pangenome graph model.
Such an effort began with the VG toolkit~\citep{Garrison:2018}.
Here we refocus it with ODGI, a compatible, but independent set of algorithms focused on visualization, interrogation, and transformation of pangenome graphs.

@@

\section{Implementation}
The ODGI toolkit builds on existing approaches to efficiently store and manipulate variation graphs \citep{Garrison:2018}.
Similar to other efficient libaries presenting the \textsc{HandleGraph} model \citep{Eizenga_2020_BX}, ODGI's implementation rests on three key properties which hold for most pangenome variation graphs:

\begin{enumerate}
\item They are relatively sparse, with low average node degree.
\item They can be sorted so that most edges go between nodes that are close together in the sort order.
\item Their embedded paths are locally similar to each other.
%We incorporate concepts first introduced in the dynamic version of the Graph BWT \citep{Siren:2020} (GBWT), a scalable implementation of the graph extension of the positional Burrows-Wheeler transform \citep{Durbin_2014}.
\end{enumerate}

These properties are used to build efficient dynamic variation graph data structures \citep{Siren:2020,Eizenga_2020_BX}.
Sparsity (1) allows us to encode edges $E$ using adjacency lists rather than matrices or hash tables.
The local linear structure of the graph (2) lets us assign node identifiers that increase along the linear components of the graph, which supports a compact storage of edges and path steps as relativistic (usually small) differences rather than absolute (always large) integer identifiers.
Path similarity (3) allows us to write local compressors that reduce the storage cost of collections of path steps.
%\FIXME{without a figure we may lose the reader here a little} -- see algorithm 1

%Here, we build on this work and focus on issues that have arisen during our recent work on the Human Pangenome.
ODGI improves on prior efforts, based on issues that arose during our work with high-quality assemblies that cover almost all parts of the genome \citep{Logsdon_2021,Nurk_2021}.
%A key design objective, motivated by experience working with graphs built from high quality eukaryotic genome assemblies, is to support graphs with very high path depth.
For example, we find that it is necessary to support graphs that have regions of very high path depth.
Such motifs can occur in collapsed repeat structures generated by ambiguous sequence homology relationships in repeats found in the centromeres and other segmental duplications.
If we cannot process such regions, we have only two options: 1) we can remove these regions, or 2) we can leave them unaligned. However, neither of these solutions allows us to investigate their biological features.
To seamlessy represent such difficult regions, we follow an approach implemented in the dynamic version of the Graph BWT (GBWT) \citep{Siren:2020}, and built a node-centric, dynamic, compressed model of the paths.
This design supports node-local modification and update of the graph, which lets us build and transform the model in parallel.

We store the graph in a vector of node structures, each of which presents a node-local view of the graph sequence, topology, and path layout.
Expressed in terms of the variation graph $V$, ODGI's core $Node$ structure includes a decoder that maps the neighbors of each node to a dense range of integers.
For a given node $i$ and neighbor $j$, the decoder itself does not store the $id$ of $j$, but rather a compact representation of the relative difference between the node ids: $\delta = Node_i.id - Node_j.id$.
This keeps even the size of the encoding small, per common variation graph property (2).
We define the edges and path steps traversing the node in terms of this alphabet of $\delta$'s.
The structures in Algorithm \ref{alg:structs} describes our encoding.

@@

Each structure contains the sequence of the node ($Node_i.sequence$), its edges in both directions ($Node_i.edges$), and a vector of path steps that describes the previous and next steps in paths that walk across the node ($Node_i.path\_steps$).
%As in the GBWT, to compress the encoding, we encode path steps using a local alphabet that maps the N neighbours of the node into the range $1\ldots N$.
%To further save space, node deltas, rather than IDs, are stored in this alphabet.
%The delta between two nodes is defined as their distance in the graph vector (i.e., the difference between the node offsets).
For efficiency, $Node_i.sequence$ is stored as a plain string, while the $edges$ and $path\_steps$ are stored using a dynamic succinct integer vector that requires $O(2nw)$ bits for the edges and $O(5nw)$ bits for the path steps, where $n$ is the number of steps on the node, and $w$ is $\approx log_2(n)$ \citep{prezza2017framework}.

To allow edit operations in parallel, each node structure includes a byte-width mutex $lock$.
All edit operations on the graph must touch at most two $Node$ structs at a time (both edge and path step representations are doubly-linked).
To avoid deadlocks, we acquire the node locks in ascending $Node.id$ order and release them in desceding order.
In addition to node-local features of the graph, we must maintain some global information.
Specifically, we record the start and end of paths, as well as a name to path id mapping in lock-free hashtables.
The use of lock-free hashtables lets us avoid a global lock when looking up path or graph metadata, which would quickly become a bottleneck during parallel operations on the graph.
By avoiding global locks, we implement many of the operations in ODGI using maximum parallelism available.
This ``atomic'' approach to edit operations on the graph is key to enable our methods to scale to the largest pangenome graphs that we can currently build.

>>>>

Pangenome graphs can be constructed by multiple sequence alignment~\citep{Lee_2002,Grasso_2004} or by transitively reducing an alignment between sequences to an equivalent, labeled sequence graph~\citep{Kehr_2014,Garrison_2019_thesis}.
Current methods to build these graphs are still under active development~\citep{Li:2020,Armstrong:2020,pggb}, but they have largely settled on a common data model, represented in the Graphical Fragment Assembly (GFA) format~\citep{GFA}).
%\FIXME{may want to move GFA standard+VG to intro section 1}. AG: it is borderline, but it seems fine to have it here too.
This standardization supports the development of a reference set of tools that operate on the pangenome graph model.
Such an effort began with the VG toolkit~\citep{Garrison:2018}.
Here we refocus it with ODGI, a compatible, but independent set of algorithms focused on visualization, interrogation, and transformation of pangenome graphs.

@@

\section{Implementation}
The ODGI toolkit builds on existing approaches to efficiently store and manipulate variation graphs~\citep{Garrison:2018}.
Similar to other efficient libaries presenting the \textsc{HandleGraph} model~\citep{Eizenga_2020_BX}, the implementation of ODGI's tools rests on three key properties which hold for most pangenome variation graphs:

\begin{enumerate}
\item They are relatively sparse, with low average node degree.
\item They can be sorted so that most edges go between nodes that are close together in the sort order.
\item Their embedded paths are locally similar to each other.
\end{enumerate}

These properties are used to build efficient dynamic variation graph data structures~\citep{Siren:2020,Eizenga_2020_BX}.
Sparsity (1) allows us to encode edges $E$ using adjacency lists rather than matrices or hash tables.
The local linear structure of the graph (2) lets us assign node identifiers that increase along the linear components of the graph, which supports a compact storage of edges and path steps as relativistic (usually small) differences rather than absolute (always large) integer identifiers.
Path similarity (3) allows us to write local compressors that reduce the storage cost of collections of path steps.
%\FIXME{without a figure we may lose the reader here a little} -- see algorithm 1

ODGI improves on prior efforts, based on issues that arose during our work with high-quality \textit{de novo} assemblies that cover almost all parts of the human genome~\citep{Logsdon_2021,Nurk_2021}.
For example, we find that it is necessary to support graphs that have regions with paths that present very high depth nodes.
Such motifs can occur in collapsed repeat structures generated by ambiguous sequence homology relationships in repeats found in the centromeres and other segmental duplications.
If we cannot process such regions, we have only two options: 1) we can remove these regions, or 2) we can leave them unaligned.
However, neither of these solutions allows us to investigate their biological features.
To seamlessly represent such difficult regions, we follow an approach implemented in the dynamic version of the Graph BWT (GBWT)~\citep{Siren:2020} and built a node-centric, dynamic, compressed model of the paths.
This design supports node-local modification and update of the graph, which lets us operate on paths in parallel.

We store the graph in a vector of node structures, each of which presents a node-local view of the graph sequence, topology, and path layout.
Expressed in terms of the variation graph $V$, ODGI's core $Node$ structure includes a decoder that maps the neighbors of each node to a dense range of integers.
For a given node $i$ and neighbor $j$, the decoder itself does not store the $id$ of $j$, but rather a compact representation of the relative difference between the node ids: $\delta = Node_i.id - Node_j.id$.
%The $\delta between two nodes is computed as their distance in the graph vector (i.e., the difference between the node offsets).
This keeps the size of the encoding small, per common variation graph property (2).
We define the edges and path steps traversing the node in terms of this alphabet of $\delta$'s.
The structures in Algorithm~\ref{alg:structs} describes our encoding.

@@

Each structure contains the sequence of the node ($Node_i.sequence$), its edges in both directions ($Node_i.edges$), and a vector of path steps that describes the previous and next steps in paths that walk across the node ($Node_i.path\_steps$).
For efficiency, $Node_i.sequence$ is stored as a plain string, while the $edges$ and $path\_steps$ are stored using a dynamic succinct integer vector that requires $O(2nw)$ bits for the edges and $O(5nw)$ bits for the path steps, where $n$ is the number of steps on the node, and $w$ is $\approx log_2(n)$ \citep{prezza2017framework}.

To allow edit operations in parallel, each node structure includes a byte-width mutex $lock$.
All edit operations on the graph must touch at most two $Node$ structs at a time (both edge and path step representations are doubly-linked).
To avoid deadlocks, we acquire the node locks in ascending $Node.id$ order and release them in desceding order.
In addition to node-local features of the graph, we must maintain some global information.
Specifically, we record the start and end of paths, as well as a name to path id mapping in lock-free hashtables.
The use of lock-free hashtables lets us avoid a global lock when looking up path or graph metadata, which would quickly become a bottleneck during parallel operations on the graph.
By avoiding global locks, we implement many of the operations in ODGI using maximum parallelism available.
This approach is key to enable our methods to scale to the largest pangenome graphs that we can currently build.
